---
title: "DATS6101 Midterm Project: Modeling Airline Passenger Satisfaction"
author: "Jehan Bugli"
date: "2023-10-15"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
   # fig_width: 14
   # fig_height: 6
  
---
```{r init, include=F}
# Install packages
#install.packages(c('ezids', 'Hmisc', 'ggplot2', 'dplyr', 'pals'))
# Load the ezids package for functions like xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
# Load package for "describe" function
library(Hmisc)
# Load ggplot2 package
library(ggplot2)
# Load dplyr package
library(dplyr)
# Load car package for VIF
library(car)
# Load corrplot package for correlation matrix visuals
library(corrplot)
# Load kableExtra package for ratings group table
library(kableExtra)
library(knitr)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, warning = FALSE)
```
<br>
<br>

## Raw Data and Filtering

```{r import_data, include = FALSE}
# Get working directory
getwd()
# Import train.csv data as a data frame
data_raw <- data.frame(read.csv("train.csv", header = TRUE))
# Get initial description of the data
str(data_raw)
header <- xkabledplyhead(data_raw, title = "Header rows for airline passenger dataset")
header


```
Responses for the ratings variables are coded as values from 1-5. However, some responses include 0, representing that the question was not applicable. Respondents that select this option for any of the ratings variables are filtered out. Fields for index number and survey ID are also removed.

<br>

The filtered dataset looks as follows:

```{r filter_ratings, echo = FALSE}
# Get unnecessary columns
drop <- c("X","id")
# Drop column names specified in vector
data <- data_raw[,!(names(data_raw) %in% drop)]

#Select ratings columns
selected_columns <- 7:20

# Check if any ratings include zeros (representing N/A)
has_zeros <- apply(data[selected_columns], 1, function(row) any(row == 0))

# Remove rows with zeros in the selected columns
data <- data[!has_zeros, ]

# Get header for filtered dataset
header <- xkabledplyhead(data, title = "Header rows for airline passenger dataset")
header
summary <- xkablesummary(data, title = "Summary of airline passenger dataset")
summary
```

<br>
<br>


## Exploratory Data Analysis (EDA)

<br>


### EDA: Ratings variables

<br>

The dataset includes a large number of different ratings variables. Most variables have a mean response value centered around 3 points, but the distribution varies for each field.

```{r ratings_histogram, fig.width = 15, fig.height = 10, echo = FALSE}

# Define ratings subset of main data frame
data_ratings <- data[7:20]

describe(data_ratings)

# Reshape the data into a long format
data_long <- tidyr::gather(data_ratings, variable, value)

colors <- c(
  "red", "blue", "green", "purple", "orange",
  "pink", "gray", "yellow", "cyan", "magenta",
  "brown", "violet", "turquoise", "pink"
)

commas <- function(x) {
  format(x, big.mark = ",")
}

# Create a clustered histogram with color-separated columns
ggplot(data_long, aes(x = value, fill = variable)) +
  geom_histogram(binwidth = 0.5, position = position_dodge(0.7), alpha = 0.7) +  # Create the histogram
  labs(x = "\nValue", y = "Frequency\n") +  # Add axis labels
  scale_fill_manual(values = colors) +  # Set custom colors
  theme_minimal() +  # Apply a minimal theme
  theme(
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines")
  ) +
  scale_y_continuous(labels = commas) + 
  labs(fill = "Ratings Variables")
```
<br>

At first glance, many of the ratings included in the histogram appear to share similar frequency distributions, sparking multicollinearity concerns. Our next step to evaluate these potential relationships was to create a correlation matrix.

<br>

```{r ratings_matrix_initial, fig.width = 15, fig.height = 15, echo = FALSE}
  
# Create a correlation matrix
cor_matrix_1 <- cor(data_ratings, method = 'spearman')

corrplot.mixed(
  cor_matrix_1,  
  lower = "number",
  upper = "circle",
  tl.pos = "lt",
  tl.cex = 1,
  number.cex = 1,
  cl.cex = 1.5,
  title="Correlation Matrix for all 5-Point Ratings", 
  cex.main = 2,
  
  mar=c(0,0,5,0) # fixes the position of title 
  
  ) 


```

<br>

We can see from the matrix that certain ratings variables have strong positive correlations with each other. If these are included in the model without adjustments, our model may suffer a loss in reliability.

<br>

In order to avoid this issue, we elected to combine ratings variables into two groups—based on the degree of correlation—and utilize average ratings from these two groups as model inputs.

<br>

| **Ratings Group 1: Pre-Flight & Wi-Fi** |
| :------: |
| In-Flight Wifi Service              |
| Departure / Arrival Time            |
| Ease of Online Booking              |
| Gate Location                       |
| Online Boarding                     |
| ------                              |  

<br>


| **Ratings Group 2: In-Flight & Baggage** |
| :------: |
| Food and Drink                     |
| Seat Comfort                       |
| In-Flight Entertainment            |
| Onboard Service                    |
| Leg Room Service                   |
| Baggage Handling                   |
| Check-In Service                   |
| In-Flight Service                  |
| Cleanliness                        |
| ------                             |


<br>

```{r ratings_combine, include = FALSE}

# Select columns for Group1
ratings_group1 <- select(data, Inflight.wifi.service, Departure.Arrival.time.convenient, Ease.of.Online.booking, Gate.location, Online.boarding)  

# Calculate the average for Group1
data$Pre_Flight_and_WiFi_Ratings <- rowMeans(ratings_group1, na.rm = TRUE)

# Select columns for Group2
ratings_group2 <- select(data, Food.and.drink, Seat.comfort, Inflight.entertainment, On.board.service, Leg.room.service, Baggage.handling, Checkin.service, Inflight.service, Cleanliness)

# Calculate the average for Group2
data$In_Flight_and_Baggage_Ratings <- rowMeans(ratings_group2, na.rm = TRUE)
summary(data)

data_ratings_combined <- data[c("Pre_Flight_and_WiFi_Ratings","In_Flight_and_Baggage_Ratings")]


```

```{r ratings_matrix_final, include = FALSE}

cor_matrix_2 <- cor(data_ratings_combined, method = 'spearman')

ratings_correlation_coeff_final <- cor_matrix_2['Pre_Flight_and_WiFi_Ratings','In_Flight_and_Baggage_Ratings']

ratings_correlation_coeff_final

```

<br>

The two consolidated ratings variables share a weak positive correlation (correlation coefficient = **`r ratings_correlation_coeff_final`**), indicating that they can be jointly included in our model without violating the collinearity assumption.

<br>


```{r ratings_final_visual, fig.width = 15, fig.height = 15, fig.align = 'center', echo = FALSE}

# Define ratings subset of main data frame
data_ratings_final <- data[c('Pre_Flight_and_WiFi_Ratings','In_Flight_and_Baggage_Ratings')]

# Transform data
data_long_final <- tidyr::gather(data_ratings_final, variable, value)

# Define custom color palette (necessary for large number of ratings)
colors <- c(
  "red", "blue", "green", "purple", "orange",
  "pink", "gray", "yellow", "cyan", "magenta",
  "brown", "violet", "turquoise", "pink"
)

commas <- function(x) {
  format(x, big.mark = ",")
}

# Create a clustered histogram with color-separated columns
ratingsgraph_final <- ggplot(data_long_final, aes(x = value, fill = variable)) +
  geom_histogram(binwidth = 0.5, position = position_dodge(1), alpha = 0.7) +  # Create the histogram
  labs(x = "\nValue", y = "Frequency\n") +  
  scale_fill_manual(values = colors) +  
  theme(
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold") 
  ) +
  scale_y_continuous(labels = commas) + 
  labs(fill = "Ratings Variables")

ratingsgraph_final + ggtitle("Histogram of aggregated ratings variables\n")
```
<br>


<br>

### EDA: Linearity with log odds

<br>
```{r encode satisfaction}
# Encode the satisfaction variable as 1/0 to use temporarily for JB analysis section, should be removed since this will be closer to the start in the full script

data$satisfaction <- ifelse(data$satisfaction == "satisfied", 1, 0)

# REMOVE IN FINAL
```
<br>
```{r discrete, fig.width = 15, fig.height = 15, fig.align = 'center', echo = FALSE}

# Calculate the discrete buckets for each variable
age_breaks <- seq(0, 90, by = 2)
dist_breaks <- seq(0, 5000, by = 20)
delay_breaks <- seq(0, 1750, by = 10)
rating_breaks <- seq(0,5,by=0.1)

# Use cut() to create breaks
data$AgeCategory <- cut(data$Age, breaks = age_breaks)
data$DistCategory <- cut(data$Flight.Distance, breaks = dist_breaks)
data$DepDelayCategory<- cut(data$Departure.Delay.in.Minutes, breaks = delay_breaks)
data$ArrDelayCategory<- cut(data$Arrival.Delay.in.Minutes, breaks = delay_breaks)
data$PreFlightCategory<-cut(data$Pre_Flight_and_WiFi_Ratings, breaks = rating_breaks)
data$InFlightCategory<-cut(data$In_Flight_and_Baggage_Ratings, breaks = rating_breaks)




```

<br>

Unlike a standard linear regression, which assumes that independent parameters have a linear relationship with the dependent variable, **logistic regression assumes that parameters have a linear relationship with the log odds**.\

Odds represent the number of favorable outcomes divided by the number of unfavorable outcomes. Put differently, if "p" represents the probability of favorable outcomes, Odds = **p/(1-p)**. Log odds take the natural log of the odds, which can be expressed as **ln(p/1-p))**.

We can use a visual test to examine whether or not this assumption holds true for continuous variables. While it is not sensible to compute log odds for individual data points, we can group continuous variables into discrete buckets to examine whether or not they would satisfy this assumption.

<br>

```{r log_odds, include = FALSE}

# Define a function to calculate the log odds for a given x value, to use with aggregate()

log_odds_calc <- function(x) {
  avg <- mean(x)
  log_odds <- log(avg / (1 - avg))
  return(log_odds)
}

# Use aggregate with the log odds function for each continuous variable of interest (with their discrete grouping)
log_odds_Age <- aggregate(satisfaction ~ AgeCategory, data, log_odds_calc)
log_odds_Dist <- aggregate(satisfaction ~ DistCategory, data, log_odds_calc)
log_odds_DepDelay <- aggregate(satisfaction ~ DepDelayCategory, data, log_odds_calc)
log_odds_ArrDelay <- aggregate(satisfaction ~ ArrDelayCategory, data, log_odds_calc)

log_odds_PreFlight <- aggregate(satisfaction ~ PreFlightCategory, data, log_odds_calc)
log_odds_InFlight <- aggregate(satisfaction ~ InFlightCategory, data, log_odds_calc)

```
<br>

```{r log_odds_cont_visual, fig.width = 15, fig.height = 15, fig.align = 'center', echo = FALSE}

# Create scatter plots for "original" continuous variables

ggplot(log_odds_Age, aes(x = AgeCategory, y = satisfaction)) +
  geom_point(size=6, color="turquoise") +
  labs(
    x = "\nAge",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Age\n"
  ) + 
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )  


ggplot(log_odds_Dist, aes(x = DistCategory, y = satisfaction)) +
  geom_point(size=6, color="blue") +
  labs(
    x = "\nDistance",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Distance\n"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )


ggplot(log_odds_DepDelay, aes(x = DepDelayCategory, y = satisfaction)) +
  geom_point(size=6, color="purple") +
  labs(
    x = "\nDeparture Delay",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Departure Delay\n"
  ) + 
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )


ggplot(log_odds_ArrDelay, aes(x = ArrDelayCategory, y = satisfaction)) +
  geom_point(size=6, color="pink") +
  labs(
    x = "\nArrival Delay",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Arrival Delay\n"
  ) +
  
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  ) 

``` 

<br>
<br>
<br>
Based on the graphs, it appears that only flight distance has a roughly linear relationship with log odds of satisfaction. Age appears to have a parabolic relationship, peaking in the middle; some sort of aggressive transformation method may be required to reach a linear relationship. Meanwhile, log odds for both delay statistics quickly disperse in both directions as they increase (likely in part due to the limited frequency of higher durations), making it difficult to conclude with certainty that a linear relationship exists.


```{r log_odds_ratings_visual, fig.width = 15, fig.height = 15, fig.align = 'center', echo = FALSE}

# Create scatter plots for aggregate ratings variables

ggplot(log_odds_PreFlight, aes(x = PreFlightCategory, y = satisfaction)) +
  geom_point(size=6, color="red") +
  labs(
    x = "\nPre-Flight & Wifi Satisfaction Rating",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Pre-Flight & Wifi Ratings\n"
  ) + 
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )  

ggplot(log_odds_InFlight, aes(x = InFlightCategory, y = satisfaction)) +
  geom_point(size=6, color="orange") +
  labs(
    x = "\nIn-Flight & Baggage Satisfaction Rating",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by In-Flight and Baggage Ratings\n"
  ) + 
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )  


```
<br>

In-flight and baggage ratings have a strikingly linear relationship with log odds; meanwhile, pre-flight and wi-fi ratings appear to have a significantly looser connection with a potential dip in log odds for average ratings. We can conclude with confidence that the in-flight aggregate fulfills the linearity assumption, while the pre-flight ratings are far more obscure in that regard. 
