---
title: "DATS6101 Midterm Project: Modeling Airline Passenger Satisfaction"
author: "Jehan Bugli"
date: "2023-10-15"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
   # fig_width: 14
   # fig_height: 6
  
---
```{r init, include=F}
# Install packages
#install.packages(c('ezids', 'Hmisc', 'ggplot2', 'dplyr', 'pals'))
# Load the ezids package for functions like xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
# Load package for "describe" function
library(Hmisc)
# Load ggplot2 package
library(ggplot2)
# Load dplyr package
library(dplyr)
# Load car package for VIF
library(car)
# Load corrplot package for correlation matrix visuals
library(corrplot)
# Load kableExtra package for ratings group table
library(kableExtra)
library(knitr)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, warning = FALSE)
```
<br>
<br>

## Raw Data and Filtering

```{r import_data, include = FALSE}
# Get working directory
getwd()
# Import train.csv data as a data frame
data_raw <- data.frame(read.csv("train.csv", header = TRUE))
# Get initial description of the data
str(data_raw)
header <- xkabledplyhead(data_raw, title = "Header rows for airline passenger dataset")
header


```
Responses for the ratings variables are coded as values from 1-5. However, some responses include 0, representing that the question was not applicable. Respondents that select this option for any of the ratings variables are filtered out. Fields for index number and survey ID are also removed.

<br>

The filtered dataset looks as follows:

```{r filter_ratings, echo = FALSE}
# Get unnecessary columns
drop <- c("X","id")
# Drop column names specified in vector
data <- data_raw[,!(names(data_raw) %in% drop)]

#Select ratings columns
selected_columns <- 7:20

# Check if any ratings include zeros (representing N/A)
has_zeros <- apply(data[selected_columns], 1, function(row) any(row == 0))

# Remove rows with zeros in the selected columns
data <- data[!has_zeros, ]

# Get header for filtered dataset
header <- xkabledplyhead(data, title = "Header rows for airline passenger dataset")
header
summary <- xkablesummary(data, title = "Summary of airline passenger dataset")
summary
```

<br>
<br>


## Exploratory Data Analysis (EDA)

<br>


### EDA: Ratings variables

<br>

The dataset includes a large number of different ratings variables. Most variables have a mean response value centered around 3 points, but the distribution varies for each field.

```{r ratings_histogram, fig.width = 15, fig.height = 10, echo = FALSE}

# Define ratings subset of main data frame
data_ratings <- data[7:20]

describe(data_ratings)

# Reshape the data into a long format
data_long <- tidyr::gather(data_ratings, variable, value)

colors <- c(
  "red", "blue", "green", "purple", "orange",
  "pink", "gray", "yellow", "cyan", "magenta",
  "brown", "violet", "turquoise", "pink"
)

commas <- function(x) {
  format(x, big.mark = ",")
}

# Create a clustered histogram with color-separated columns
ggplot(data_long, aes(x = value, fill = variable)) +
  geom_histogram(binwidth = 0.5, position = position_dodge(0.7), alpha = 0.7) +  # Create the histogram
  labs(x = "\nValue", y = "Frequency\n") +  # Add axis labels
  scale_fill_manual(values = colors) +  # Set custom colors
  theme_minimal() +  # Apply a minimal theme
  theme(
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines")
  ) +
  scale_y_continuous(labels = commas) + 
  labs(fill = "Ratings Variables")
```
<br>

At first glance, many of the ratings included in the histogram appear to share similar frequency distributions, sparking multicollinearity concerns. Our next step to evaluate these potential relationships was to create a correlation matrix.

<br>

```{r ratings_matrix_initial, fig.width = 15, fig.height = 15, echo = FALSE}
  
# Create a correlation matrix
cor_matrix_1 <- cor(data_ratings, method = 'spearman')

corrplot.mixed(
  cor_matrix_1,  
  lower = "number",
  upper = "circle",
  tl.pos = "lt",
  tl.cex = 1,
  number.cex = 1,
  cl.cex = 1.5,
  title="Correlation Matrix for all 5-Point Ratings", 
  cex.main = 2,
  
  mar=c(0,0,5,0) # fixes the position of title 
  
  ) 


```

<br>

We can see from the matrix that certain ratings variables have strong positive correlations with each other. If these are included in the model without adjustments, our model may suffer a loss in reliability.

<br>

In order to avoid this issue, we elected to combine ratings variables into two groups—based on the degree of correlation—and utilize average ratings from these two groups as model inputs.

<br>

| **Ratings Group 1: Pre-Flight & Wi-Fi** |
| :------: |
| In-Flight Wifi Service              |
| Departure / Arrival Time            |
| Ease of Online Booking              |
| Gate Location                       |
| Online Boarding                     |
| ------                              |  

<br>


| **Ratings Group 2: In-Flight & Baggage** |
| :------: |
| Food and Drink                     |
| Seat Comfort                       |
| In-Flight Entertainment            |
| Onboard Service                    |
| Leg Room Service                   |
| Baggage Handling                   |
| Check-In Service                   |
| In-Flight Service                  |
| Cleanliness                        |
| ------                             |


<br>

```{r ratings_combine, include = FALSE}

# Select columns for Group1
ratings_group1 <- select(data, Inflight.wifi.service, Departure.Arrival.time.convenient, Ease.of.Online.booking, Gate.location, Online.boarding)  

# Calculate the average for Group1
data$Pre_Flight_and_WiFi_Ratings <- rowMeans(ratings_group1, na.rm = TRUE)

# Select columns for Group2
ratings_group2 <- select(data, Food.and.drink, Seat.comfort, Inflight.entertainment, On.board.service, Leg.room.service, Baggage.handling, Checkin.service, Inflight.service, Cleanliness)

# Calculate the average for Group2
data$In_Flight_and_Baggage_Ratings <- rowMeans(ratings_group2, na.rm = TRUE)
summary(data)

data_ratings_combined <- data[c("Pre_Flight_and_WiFi_Ratings","In_Flight_and_Baggage_Ratings")]


```

```{r ratings_matrix_final, include = FALSE}

cor_matrix_2 <- cor(data_ratings_combined, method = 'spearman')

ratings_correlation_coeff_final <- cor_matrix_2['Pre_Flight_and_WiFi_Ratings','In_Flight_and_Baggage_Ratings']

ratings_correlation_coeff_final

```

The two consolidated ratings variables share a weak positive correlation (correlation coefficient = **`r ratings_correlation_coeff_final`**), indicating that they can be jointly included in our model without violating the collinearity assumption.


```{r ratings_matrix_final_visual, fig.width = 15, fig.height = 15, fig.align = 'center', include = FALSE}
  
# Create a correlation matrix with the combined ratings fields


corrplot.mixed(
  cor_matrix_2,  
  lower = "number",
  upper = "circle",
  tl.pos = "d",
  tl.cex = 1,
  number.cex = 1,
  cl.cex = 1.5,
  title="Correlation Matrix for consolidated 5-Point Ratings", 
  cex.main = 2,
  
  mar=c(0,0,5,0) # fixes the position of title 
  
  ) 

#title("Correlation Matrix for consolidated 5-Point Ratings", line = 1, cex.main = 2)



```

