---
title: "Decisiontreemodel"
author: "Parv Bhargava"
date: "2023-12-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(rpart)
library(rpart.plot)
library(caret)
library(broom)
library(tidyverse)
library(MASS)
library(ROCR)

```

```{r}
Train <- read.csv("data\\train.csv")
Test <- read.csv("data\\test.csv") 
```
**Column Datatype Changes - Testing Data**: Conversion of certain columns to factors based on their ordinal nature.
```{r}

Test$Inflight.wifi.service = as.factor(Test$Inflight.wifi.service)
Test$Departure.Arrival.time.convenient = as.factor(Test$Departure.Arrival.time.convenient)
Test$Ease.of.Online.booking = as.factor(Test$Ease.of.Online.booking) 
Test$Gate.location = as.factor(Test$Gate.location)
Test$Food.and.drink = as.factor(Test$Food.and.drink)
Test$Online.boarding = as.factor(Test$Online.boarding)
Test$Seat.comfort = as.factor(Test$Seat.comfort)
Test$Inflight.entertainment = as.factor(Test$Inflight.entertainment)
Test$On.board.service = as.factor(Test$On.board.service)
Test$Leg.room.service = as.factor(Test$Leg.room.service)
Test$Baggage.handling = as.factor(Test$Baggage.handling)
Test$Checkin.service = as.factor(Test$Checkin.service)
Test$Inflight.service = as.factor(Test$Inflight.service)
Test$Cleanliness = as.factor(Test$Cleanliness)
```
**Column Datatype Changes - Training Data**: Similar data type conversions for training data.
```{r}
#Column datatype Changes - Training Data - As Columns has ordinal its better to convert into factor

Train$Inflight.wifi.service = as.factor(Train$Inflight.wifi.service)
Train$Departure.Arrival.time.convenient = as.factor(Train$Departure.Arrival.time.convenient)
Train$Ease.of.Online.booking = as.factor(Train$Ease.of.Online.booking) 
Train$Gate.location = as.factor(Train$Gate.location)
Train$Food.and.drink = as.factor(Train$Food.and.drink)
Train$Online.boarding = as.factor(Train$Online.boarding)
Train$Seat.comfort = as.factor(Train$Seat.comfort)
Train$Inflight.entertainment = as.factor(Train$Inflight.entertainment)
Train$On.board.service = as.factor(Train$On.board.service)
Train$Leg.room.service = as.factor(Train$Leg.room.service)
Train$Baggage.handling = as.factor(Train$Baggage.handling)
Train$Checkin.service = as.factor(Train$Checkin.service)
Train$Inflight.service = as.factor(Train$Inflight.service)
Train$Cleanliness = as.factor(Train$Cleanliness)
```
**Replacing NA Values in Training Data - Arrival Delay in Minutes**: Handling missing values in the training dataset.
```{r}
#Replacing NA Values in Training Data - Arrival Delay in Minutes
NA.position <- which(is.na(Train$Arrival.Delay.in.Minutes))
Train$Arrival.Delay.in.Minutes[NA.position] = mean(Train$Arrival.Delay.in.Minutes, na.rm = TRUE)
```
**Replacing NA Values in Testing Data - Arrival Delay in Minutes**: Handling missing values in the testing dataset.
```{r}
#Replacing NA Values in Testing Data - Arrival Delay in Minutes
NA.position1 <- which(is.na(Test$Arrival.Delay.in.Minutes))
Test$Arrival.Delay.in.Minutes[NA.position1] = mean(Test$Arrival.Delay.in.Minutes, na.rm = TRUE)
```

```{r pressure, echo=FALSE}
tree = rpart(satisfaction ~ Gender + Customer.Type + Age + 
               Type.of.Travel + Class + Flight.Distance + Inflight.wifi.service + 
               Departure.Arrival.time.convenient + Ease.of.Online.booking + 
               Gate.location + Food.and.drink + Online.boarding + Seat.comfort +
               Inflight.entertainment + On.board.service + Leg.room.service +
               Baggage.handling + Checkin.service + Inflight.service +
               Cleanliness + Departure.Delay.in.Minutes + Arrival.Delay.in.Minutes , 
             data = Train, method = 'class', minbucket=25)
```
**Analyzing the Importance of Variable Using the Variable Importance Plot**: Analysis to identify significant variables.
```{r}
#Analyzing the Importance of variable using the Variable Importance Plot
varImp(tree)
```
**Re-running the Model with Significant Variables**: Building the model with chosen variables.
```{r}
#Re-running the model with significant variables
tree1 = rpart(satisfaction ~  Age + Type.of.Travel + Class + Inflight.wifi.service + 
                Ease.of.Online.booking + Online.boarding + Seat.comfort +
               Inflight.entertainment + On.board.service + Leg.room.service +
               Baggage.handling + Checkin.service + Inflight.service +
               Cleanliness + Arrival.Delay.in.Minutes,data = Train, 
             method = 'class', minbucket=25)
```
**Visualizing the Decision Tree**: Displaying the structure of the decision tree.
```{r}
#Visualizaing the Decision tree
prp(tree1)
```
**Define Cross-validation Experiment**: Setting up the cross-validation process.
```{r}
# Define cross-validation experiment
numFolds = trainControl( method = "cv", number = 10 )
cpGrid = expand.grid( .cp = seq(0.01,0.5,0.01))
```
**Perform the Cross-validation**: Executing the cross-validation.
```{r}
# Perform the cross validation
train(satisfaction ~  Age + Type.of.Travel + Class + Inflight.wifi.service + 
        Ease.of.Online.booking + Online.boarding + Seat.comfort +
        Inflight.entertainment + On.board.service + Leg.room.service +
        Baggage.handling + Checkin.service + Inflight.service +
        Cleanliness + Arrival.Delay.in.Minutes, 
      data = Train, method = "rpart", trControl = numFolds, tuneGrid = cpGrid )
```
**Predicting the Values on the Test Data**: Applying the model to the test data.
```{r}
tree2 = rpart(satisfaction ~  Age + Type.of.Travel + Class + Inflight.wifi.service + 
                Ease.of.Online.booking + Online.boarding + Seat.comfort +
                Inflight.entertainment + On.board.service + Leg.room.service +
                Baggage.handling + Checkin.service + Inflight.service +
                Cleanliness + Arrival.Delay.in.Minutes, 
              data = Train, method="class", cp = 0.01)
```

```{r}
#Predicting the Values on the Test data
PredictROC = predict(tree1, newdata = Test)
```

**Plotting the ROC Curve**: Visualizing the Receiver Operating Characteristic curve.
```{r}
#Plotting the ROC Curve
pred = prediction(PredictROC[,2], Test$satisfaction)
perf = performance(pred, "tpr", "fpr")
plot(perf, colorize = TRUE, print.cutoffs.at=seq(0,1,by=0.1),text.adj = c(-0.2,1.7))
```
**Confusion Matrix Table to Find Accuracy**: Evaluating the model's accuracy.
```{r}
#Confusion Matrix table to find accuracy
#From the ROC Curve, we found 0.7 is the optimum threshold value for Cut-off.
table(Test$satisfaction, PredictROC[,2] > 0.7)
```
**Optimum Threshold Value for Cut-off from ROC Curve**: Determining the best threshold value.

**Calculating Accuracy**: Detailed accuracy calculation.
```{r}
#Calculating Accuracy
Accuracy_avg_Tree = (14003+8484)/(14003+8484+570+2919)
Accuracy_avg_Tree
```
**Calculating Sensitivity or Recall Value**: Calculating the recall metric.
```{r}
#Calculating Sensitivity or Recall value
Recall = (8484)/(8484+2919)
Recall
```
**Calculating Precision Value**: Computing the precision of the model.
```{r}
#Calculating Precision Value
Precision = (8484)/(8484+570)
Precision
```
**Calculating F-Measure**: Determining the F-Measure for model evaluation.
```{r}
#Calculating F-Measure
F.measure = (2*Recall*Precision)/(Recall+Precision)
F.measure
```
**Calculating Specificity**: Assessing the specificity metric.
```{r}
#Calculating Specificity
Specificity = (14003)/(14003+570)
Specificity
```
**Testing Data AUC-ROC (Area Under the Curve - Receiver Operator Characteristics)
```{r}
#Testing Data AUC-ROC(Area Under the Curve - Receiver operator Characteristics) value
AUC = as.numeric(performance(pred, "auc")@y.values)
AUC
```

