---
title: "An Exploratory Data Analysis on Airline Customer Satisfaction"
author: "Parv Bhargava"
date: "2023-10-20"
output: html_document
---


```{r init, include=FALSE}
library(psych)
library(readr)
library(forcats)
library(gridExtra)
library(RColorBrewer)
library(usdm)
# Load the ezids package for functions like xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
# Load package for "describe" function
library(Hmisc)
# Load ggplot2 package
library(ggplot2)
# Load dplyr package
library(dplyr)
# Load car package for VIF
library(car)
# Load corrplot package for correlation matrix visuals
library(corrplot)
# Load kableExtra package for ratings group table
library(kableExtra)
library(knitr)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, warning = FALSE)
```

## Introduction


### - Briefly introduce the data and its context.


### - Mention the objective of the analysis: To explore the data and demonstrate why linear regression may not be the best model for this specific case.

## Load the Data

Import the data into R by using `read.csv()` function.

```{r data}
data <- read_csv("archive/train.csv")
xkabledplyhead(data)
```

### Check its structure, dimensions.
```{r 'structure and dimension of the data'}
str(data)
dim(data)
```

**Data Preprocessing**
```{r data preprocessing}
# Check for missing values
missing_data <- data %>%
  summarise_all(~ sum(is.na(.)))

# Check for duplicated rows
duplicate_rows <- data %>%
  summarise(n_duplicates = sum(duplicated(.)))

# Remove duplicated rows if necessary
# data <- data[!duplicated(data), ]
missing_data
duplicate_rows

missing_data <- sum(is.na(data$`Arrival Delay in Minutes`))

data$`Arrival Delay in Minutes`[is.na(data$`Arrival Delay in Minutes`)] <- median(data$`Arrival Delay in Minutes`, na.rm = TRUE)

#duplicate
sum(duplicated(data))
missing_data
```

```{r histograms_categorical, fig.width=15, fig.height=15}

categorical_vars_ggplot <- c('Gender', '`Customer Type`', '`Type of Travel`', 'Class', 'satisfaction')

plot_list <- list()

for (cat_var in categorical_vars_ggplot) {
  plot_obj <- ggplot(data, aes_string(x = cat_var, fill = cat_var)) + 
    geom_bar() +
    geom_text(stat='count', aes_string(label='..count..', y='..count..'), vjust=-0.5) +
    labs(title = paste("Distribution of", cat_var), x = cat_var, y = "Count") +
    scale_fill_brewer(palette="Set3") + 
    theme_minimal() +  
    theme(legend.position="none")
  
  plot_list[[cat_var]] <- plot_obj
}

grid.arrange(grobs = plot_list, ncol = 2)


```

```{r variables}

numerical_vars_ggplot <- c('Age', '`Flight Distance`', '`Inflight wifi service`', '`Departure/Arrival time convenient`', '`Ease of Online booking`', '`Gate location`', '`Food and drink`', '`Online boarding`', '`Seat comfort`', '`Inflight entertainment`', '`On-board service`', '`Leg room service`', '`Baggage handling`', '`Checkin service`', '`Inflight service`', 'Cleanliness', '`Departure Delay in Minutes`', '`Arrival Delay in Minutes`')

continuous_vars <- c('Age', '`Flight Distance`', '`Departure Delay in Minutes`', '`Arrival Delay in Minutes`')
```

```{r continous_variables, fig.width=10, fig.height=10}

continuous_plots <- list()

for (num_var in continuous_vars) {
    plot_obj <- ggplot(data, aes_string(x = num_var)) + 
      geom_histogram(bins = 40, fill = "skyblue", color = "black", alpha = 0.7) + 
      labs(title = paste("Distribution of", num_var), x = num_var, y = "Frequency") +
      theme_minimal()
    
    continuous_plots <- append(continuous_plots, list(plot_obj))
}

if (length(continuous_plots) > 0) {
  grid.arrange(grobs = continuous_plots, ncol = 2)
}

```
```{r rating_plots, fig.width=50, fig.height=100}

rating_vars <- c('`Inflight wifi service`', '`Departure/Arrival time convenient`', '`Ease of Online booking`', '`Gate location`', '`Food and drink`', '`Online boarding`', '`Seat comfort`', '`Inflight entertainment`', '`On-board service`', '`Leg room service`', '`Baggage handling`', '`Checkin service`', '`Inflight service`', 'Cleanliness')

rating_plots <- list()

for (rate_var in rating_vars) {
    plot_obj <- ggplot(data, aes_string(x = rate_var)) + 
      geom_bar(aes(fill = get(rate_var))) +
      geom_text(stat='count', aes_string(label='..count..', y='..count..'), vjust=0, size = 20) +
      labs(title = paste("Distribution of", rate_var), x = rate_var, y = "Frequency")+
      scale_fill_brewer(palette="Set3") + 
      theme_minimal() +  
      theme(legend.position="none",
            
        plot.title = element_text(size = 54, face = "bold"),
        axis.title.x = element_text(size = 52, face = "bold"),  
        axis.title.y = element_text(size = 52, face = "bold"),
        axis.text.x = element_text(size = 50),
        axis.text.y = element_text(size = 50))
    
    rating_plots <- append(rating_plots, list(plot_obj))
}

if (length(rating_plots) > 0) {
  grid.arrange(grobs = rating_plots, ncol = 2)
}
```

## BoxPlots Related to the Target Variable(Satisfaction)

#### we are exploring how different numerical features interact with the target variable (satisfaction). This will help in understanding how different values of a feature relate to customer satisfaction.

```{r box_plot, fig.width=10, fig.height=10}

library(gridExtra)

histogram_vars <- c('Age', '`Flight Distance`', '`Departure Delay in Minutes`', '`Arrival Delay in Minutes`')

boxplot_list <- list()

for (num_var in histogram_vars) {
  plot_obj <- ggplot(data, aes_string(x = "satisfaction", y = num_var)) + 
      geom_boxplot(fill = "skyblue", color = "black", alpha = 0.7) +
      labs(title = paste("Box plot", num_var), 
           x = "Satisfaction", y = num_var) +
      theme_minimal() +
      theme(axis.text = element_text(size=10),
            axis.title = element_text(size=10, face="bold"),
            plot.title = element_text(hjust = 0.3))
  
  boxplot_list <- append(boxplot_list, list(plot_obj))
}

grid.arrange(grobs = boxplot_list, ncol = 2)
```

#### Age: Satisfied customers tend to be older than neutral or dissatisfied customers.

#### Flight Distance: Satisfied customers, on average, seem to travel longer distances.

#### Departure Delay in Minutes: Flights with higher departure delays tend to have more neutral or dissatisfied customers.

#### Arrival Delay in Minutes: Similarly, flights with higher arrival delays tend to have more neutral or dissatisfied customers.

```{r target_variable_histograms}

categorical_features <- c('Gender', '`Customer Type`', '`Type of Travel`', 'Class')

par(mfrow=c(2,2), mar=c(4,4,2,2))

for (feature in categorical_features) {
  p <- ggplot(data, aes_string(x=feature, fill='satisfaction')) +
    geom_bar(position="dodge") +
    geom_text(stat='count', aes(label=..count..), vjust=-0.5, position=position_dodge(width=0.9)) +
    labs(title = paste("Distribution of", gsub("`", "", feature), "by Satisfaction"), x = gsub("`", "", feature), y = "Count") +
    scale_fill_brewer(palette="Set3") +
    theme_minimal() +
    theme(legend.position="top")
  print(p)
}

```

#### Observtions

***
Gender: Both genders have a fairly similar distribution of satisfaction levels.
Customer Type: Loyal customers tend to be more satisfied than disloyal ones.
Type of Travel: Passengers traveling for business purposes are generally more satisfied than those traveling for personal reasons.
Class: Business class passengers are noticeably more satisfied than those in Eco or Eco Plus.

***

## Distribution of numerical features by satisfaction - KDE (Kernel Density Estimation)

```{r}

numerical_features <- c('Age', '`Flight Distance`', '`Departure Delay in Minutes`', '`Arrival Delay in Minutes`')

par(mfrow=c(2,2), mar=c(4,4,2,2))

for (feature in numerical_features) {
  p <- ggplot(data, aes_string(x=feature, fill='satisfaction')) +
    geom_density(alpha=0.5, position="identity") +
    labs(title = paste("Distribution of", feature, "by Satisfaction"), x = feature, y = "Density") +
    scale_fill_manual(values=c("satisfied"="green", "neutral or dissatisfied"="red")) +
    theme_minimal() +
    theme(legend.position="top")
  print(p)
}


```


#### Observations


***

Age: Younger passengers tend to be more neutral or dissatisfied, while older passengers lean more towards satisfaction.

Flight Distance: Passengers traveling shorter distances seem to be more neutral or dissatisfied compared to those traveling longer distances.

Departure Delay in Minutes: Although the distributions overlap considerably, there's a slightly higher density of neutral or dissatisfied passengers with longer departure delays.

Arrival Delay in Minutes: Similar to departure delays, passengers with longer arrival delays tend to be more neutral or dissatisfied.


***

## Visualize the relationship between Age and Flight Distance colored by satisfaction.


```{r scatter_plots}

# just 5000 rows
data_sample <- data[sample(nrow(data), 5000), ]

# Scatter plot
ggplot(data_sample, aes(x=Age, y=`Flight Distance`, color=satisfaction)) +
  geom_point(alpha=0.7) +
  scale_color_manual(values=c("neutral or dissatisfied"="red", "satisfied"="green")) +
  labs(title="Relationship between Age and Flight Distance by Satisfaction") +
  theme_minimal()
```


***

There's a spread of satisfied and neutral/dissatisfied customers across various ages and flight distances.

Older passengers who travel longer distances seem to be predominantly satisfied.

Younger passengers, especially those traveling shorter distances, display a mix of satisfaction levels.

***



```{r}


categorical_vars <- c('Gender', 'Customer Type', 'Type of Travel', 'Class')

perform_chi2_test <- function(feature) {
  contingency_table <- table(data[[feature]], data$satisfaction)
  chi2_test_result <- chisq.test(contingency_table)
  return(chi2_test_result$p.value)
}

chi2_p_values <- sapply(categorical_vars, perform_chi2_test)

names(chi2_p_values) <- categorical_vars
chi2_p_values



```


***

For all these variables, the p-values are extremely small, indicating that there is a statistically significant association between the categorical variable and the target variable satisfaction.


Given the extremely small p-values for all the categorical variables, we can reject the null hypothesis
 for each of them. This implies that there's a statistically significant association between each of these categorical variables (Gender, Customer Type, Type of Travel, and Class) and passenger satisfaction.

In simpler terms, the likelihood of a passenger being satisfied (or not) is not independent of their gender, customer type, type of travel, or class. Each of these factors plays a role in determining their satisfaction level.

***

## T-test 

Null Hypothesis : There is no difference in the means of "Arrival Delay in Minutes" between the two satisfaction groups (satisfied and neutral or dissatisfied).

alternative hypothesis : There is a significant difference in the means of "Arrival Delay in Minutes" between the two satisfaction groups.


#### We will perform a two-sample t-test to determine if the means of "Arrival Delay in Minutes" for the two satisfaction groups are statistically different.

```{r}

group1_arrival_delay <- data$`Arrival Delay in Minutes`[data$satisfaction == "satisfied"]
group2_arrival_delay <- data$`Arrival Delay in Minutes`[data$satisfaction == "neutral or dissatisfied"]

t_test_result_arrival_delay <- t.test(group1_arrival_delay, group2_arrival_delay)


t_test_result_arrival_delay



```

#### Given the very low p-value, we can reject the null hypothesis. This means there is a statistically significant difference in the means of "Arrival Delay in Minutes" between passengers who are satisfied and those who are neutral or dissatisfied. Specifically, passengers who are satisfied seem to experience, on average, a shorter arrival delay compared to those who are neutral or dissatisfied.

```{r data2load, echo = TRUE}
# Import train.csv data as a data frame
data_raw <- data.frame(read.csv("archive/train.csv", header = TRUE))
```

```{r filter_ratings, echo = FALSE}
# Get unnecessary columns
drop <- c("X","id")
# Drop column names specified in vector
data <- data_raw[,!(names(data_raw) %in% drop)]

#Select ratings columns
selected_columns <- 7:20

# Check if any ratings include zeros (representing N/A)
has_zeros <- apply(data[selected_columns], 1, function(row) any(row == 0))

# Remove rows with zeros in the selected columns
data <- data[!has_zeros, ]

# Get header for filtered data
header <- xkabledplyhead(data, title = "Header rows for airline passenger data")
header
summary <- xkablesummary(data, title = "Summary of airline passenger data")
summary
```

***Hypothesis Testing***
1. **Hypothesis**: The satisfaction level is independent of the gender of the passengers.
   - Null Hypothesis (H0): Satisfaction and gender are independent.
   - Alternate Hypothesis (H1): Satisfaction and gender are dependent.
   
   
```{r gender_hpo}
   contingency_table <- table(data$Gender, data$satisfaction)
   chi_sq_test <- chisq.test(contingency_table)
   chi_sq_test
```

2. **Hypothesis**: The type of travel is associated with passenger satisfaction.
   - Null Hypothesis (H0): The type of travel and satisfaction are independent.
   - Alternate Hypothesis (H1): The type of travel and satisfaction are dependent.
   
```{r tot_hpo}
   contingency_table <- table(data$Type.of.Travel, data$satisfaction)
   chi_sq_test <- chisq.test(contingency_table)
   chi_sq_test
```

3. **Hypothesis**: Customer type affects satisfaction levels.
   - Null Hypothesis (H0): Customer type and satisfaction are independent.
   - Alternate Hypothesis (H1): Customer type and satisfaction are dependent.
   
```{r ct_hpo}
   contingency_table <- table(data$Customer.Type, data$satisfaction)
   chi_sq_test <- chisq.test(contingency_table)
   chi_sq_test
```

4. **Hypothesis**: There is a relationship between class and passenger satisfaction.
   - Null Hypothesis (H0): Class and satisfaction are independent.
   - Alternate Hypothesis (H1): Class and satisfaction are dependent.
   
```{r class}
   contingency_table <- table(data$Class, data$satisfaction)
   chi_sq_test <- chisq.test(contingency_table)
   chi_sq_test
```

5. **Hypothesis**: Inflight WiFi service has an impact on passenger satisfaction.
   - Null Hypothesis (H0): Inflight WiFi service and satisfaction are independent.
   - Alternate Hypothesis (H1): Inflight WiFi service and satisfaction are dependent.
   
```{r wifi_hyp0}
   contingency_table <- table(data$Inflight.wifi.service, data$satisfaction)
   chi_sq_test <- chisq.test(contingency_table)
   chi_sq_test
```

6. **Hypothesis**: Gender and the convenience of departure/arrival times are independent.
   - Null Hypothesis (H0): Gender and departure/arrival time convenience are independent.
   - Alternate Hypothesis (H1): Gender and departure/arrival time convenience are dependent.
   
```{r dat_hyp0}
   contingency_table <- table(data$Gender, data$Departure.Arrival.time.convenient)
   chi_sq_test <- chisq.test(contingency_table)
   chi_sq_test
```

7. **Hypothesis**: Ease of online booking and satisfaction are related.
   - Null Hypothesis (H0): Ease of online booking and satisfaction are independent.
   - Alternate Hypothesis (H1): Ease of online booking and satisfaction are dependent.
   
```{r olbokking_hyp0}
   contingency_table <- table(data$Ease.of.Online.booking, data$satisfaction)
   chi_sq_test <- chisq.test(contingency_table)
   chi_sq_test
```

8. **Hypothesis**: The location of the gate affects passenger satisfaction.
   - Null Hypothesis (H0): Gate location and satisfaction are independent.
   - Alternate Hypothesis (H1): Gate location and satisfaction are dependent.
   
```{r gate_hypo}
   contingency_table <- table(data$Gate.location, data$satisfaction)
   chi_sq_test <- chisq.test(contingency_table)
   chi_sq_test
```

9. **Hypothesis**: Food and drink service quality impacts passenger satisfaction.
   - Null Hypothesis (H0): Food and drink service and satisfaction are independent.
   - Alternate Hypothesis (H1): Food and drink service and satisfaction are dependent.
   
```{r serv_hyp0}
   contingency_table <- table(data$Food.and.drink, data$satisfaction)
   chi_sq_test <- chisq.test(contingency_table)
   chi_sq_test
```

10. **Hypothesis**: Online boarding experience influences passenger satisfaction.
    - Null Hypothesis (H0): Online boarding and satisfaction are independent.
    - Alternate Hypothesis (H1): Online boarding and satisfaction are dependent.
    
```{r onlinebord_hypo}
    contingency_table <- table(data$Online.boarding, data$satisfaction)
    chi_sq_test <- chisq.test(contingency_table)
    chi_sq_test
```

### Hypothesis Tests for Numeric Variables:

1. **Hypothesis**: The mean age of satisfied passengers is equal to the mean age of dissatisfied passengers.
   - Null Hypothesis (H0): μ_age_satisfied = μ_age_dissatisfied
   - Alternate Hypothesis (H1): μ_age_satisfied ≠ μ_age_dissatisfied

```{r age_rel}
   t_test <- t.test(data$Age ~ data$satisfaction)
   t_test
```

2. **Hypothesis**: The mean flight distance for business class passengers is different from that of economy class passengers.
   - Null Hypothesis (H0): μ_distance_business = μ_distance_economy
   - Alternate Hypothesis (H1): μ_distance_business ≠ μ_distance_economy

```{r distance_hypo_anova}
# Perform one-way ANOVA to compare means across different classes
result <- aov(Flight.Distance ~ Class, data = data)
# Check the ANOVA table
summary(result)
```

3. **Hypothesis**: Departure delay minutes for satisfied passengers are less than departure delay minutes for dissatisfied passengers.
   - Null Hypothesis (H0): μ_departure_delay_satisfied ≥ μ_departure_delay_dissatisfied
   - Alternate Hypothesis (H1): μ_departure_delay_satisfied < μ_departure_delay_dissatisfied

```{r ddl_hypo}
   t_test <- t.test(data$Departure.Delay.in.Minutes ~ data$satisfaction, alternative = "less")
   t_test
```


***Correlation Analysis***
Correlation Matrix

Let's take a look at the correlation matrix of the data. This will be done in 2 parts:

1) Numerical variables

2) Categorical variables

Numerical variable Correlation Plot

```{r}
data_cor <- cor(subset(data,select = c(Age, Flight.Distance,Departure.Delay.in.Minutes,Arrival.Delay.in.Minutes)))
summary(data_cor)
options(repr.plot.width = 14, repr.plot.height = 8)
corrplot(data_cor, na.label = " ", method="color", tl.col = "black", tl.cex = 1)
```



```{r fig.width=10, fig.height=10}

categorical_fields <- c("Inflight.wifi.service", "Departure.Arrival.time.convenient", "Ease.of.Online.booking",
  "Gate.location", "Food.and.drink", "Online.boarding", "Seat.comfort",
  "Inflight.entertainment", "On.board.service", "Leg.room.service", "Baggage.handling",
  "Checkin.service", "Inflight.service", "Cleanliness")

correlation_matrix <- cor(data[categorical_fields])
options(repr.plot.width=100, repr.plot.height=80)
corrplot(correlation_matrix, method = "circle", type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 90, addCoef.col = "black", number.cex = 0.9, 
         cl.cex = 0.9)
```

Responses for the ratings variables are coded as values from 1-5. However, some responses include 0, representing that the question was not applicable. Respondents that select this option for any of the ratings variables are filtered out. Fields for index number and survey ID are also removed.

The data includes a large number of different ratings variables. Most variables have a mean response value centered around 3 points, but the distribution varies for each field.

## Multicollinearity Test
One of the essential steps in data analysis is assessing multicollinearity among independent variables. Multicollinearity occurs when predictor variables are highly correlated with each other, which can impact the reliability of regression models. To address this, we perform a VIF (Variance Inflation Factor) test.

```{r VIF}

# Select the relevant independent variables for the VIF analysis (e.g., numeric variables)
independent_vars <- data[, c(
  "Flight.Distance",
  "Inflight.wifi.service",
  "Departure.Arrival.time.convenient",
  "Ease.of.Online.booking",
  "Gate.location",
  "Food.and.drink",
  "Online.boarding",
  "Seat.comfort",
  "Inflight.entertainment",
  "On.board.service",
  "Leg.room.service",
  "Baggage.handling",
  "Checkin.service",
  "Inflight.service",
  "Cleanliness",
  "Departure.Delay.in.Minutes",
  "Arrival.Delay.in.Minutes"
)]

# Calculate VIF using the 'usdm' library
vif_values <- usdm::vif(independent_vars)

# Create a data frame to display VIF values and variables
vif_summary <- data.frame(Variable = names(independent_vars), VIF = vif_values)

# Print the VIF summary
print(vif_summary)
```

```{r VIF_Correlation_Heatmap, fig.width=10, fig.height=8}
correlation_matrix <- cor(independent_vars)
# Create a heatmap of the correlation matrix with larger dimensions
corrplot(correlation_matrix, method = "color")
```

## Interpretation and Conclusion
The VIF test results indicate the presence of multicollinearity among some independent variables. Multicollinearity can affect the reliability of regression models and the interpretation of variable contributions. In our dataset, "Departure.Delay.in.Minutes" and "Arrival.Delay.in.Minutes" have high VIF values, suggesting strong multicollinearity. This implies that these two variables may be highly correlated with other predictors in our dataset.

To address multicollinearity, we may consider:

  1. Removing one of the highly correlated variables.
  2. Using dimensionality reduction techniques to create uncorrelated predictors.
  3. Re-evaluating the model's predictors based on the context of the analysis and research goals.

It's essential to address multicollinearity to improve the quality of our predictive models and better understand the impact of various factors on passenger satisfaction.

```{r ratings_histogram, fig.width = 15, fig.height = 10, echo = FALSE}

# Define ratings subset of main data frame
data_ratings <- data[7:20]

describe(data_ratings)

# Reshape the data into a long format
data_long <- tidyr::gather(data_ratings, variable, value)

colors <- c(
  "red", "blue", "green", "purple", "orange",
  "pink", "gray", "yellow", "cyan", "magenta",
  "brown", "violet", "turquoise", "pink"
)

commas <- function(x) {
  format(x, big.mark = ",")
}

# Create a clustered histogram with color-separated columns
ggplot(data_long, aes(x = value, fill = variable)) +
  geom_histogram(binwidth = 0.5, position = position_dodge(0.7), alpha = 0.7) +  # Create the histogram
  labs(x = "\nValue", y = "Frequency\n") +  # Add axis labels
  scale_fill_manual(values = colors) +  # Set custom colors
  theme_minimal() +  # Apply a minimal theme
  theme(
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines")
  ) +
  scale_y_continuous(labels = commas) + 
  labs(fill = "Ratings Variables")
```
<br>

At first glance, many of the ratings included in the histogram appear to share similar frequency distributions, sparking multicollinearity concerns. Our next step to evaluate these potential relationships was to create a correlation matrix.

<br>

```{r ratings_matrix_initial, fig.width = 15, fig.height = 15, echo = FALSE}
  
# Create a correlation matrix
cor_matrix_1 <- cor(data_ratings, method = 'spearman')

corrplot.mixed(
  cor_matrix_1,  
  lower = "number",
  upper = "circle",
  tl.pos = "lt",
  tl.cex = 1,
  number.cex = 1,
  cl.cex = 1.5,
  title="Correlation Matrix for all 5-Point Ratings", 
  cex.main = 2,
  
  mar=c(0,0,5,0) # fixes the position of title 
  
  ) 


```

<br>

We can see from the matrix that certain ratings variables have strong positive correlations with each other. If these are included in the model without adjustments, our model may suffer a loss in reliability.

<br>

In order to avoid this issue, we elected to combine ratings variables into two groups—based on the degree of correlation—and utilize average ratings from these two groups as model inputs.

<br>

| **Ratings Group 1: Pre-Flight & Wi-Fi** |
| :------: |
| In-Flight Wifi Service              |
| Departure / Arrival Time            |
| Ease of Online Booking              |
| Gate Location                       |
| Online Boarding                     |
| ------                              |  

<br>


| **Ratings Group 2: In-Flight & Baggage** |
| :------: |
| Food and Drink                     |
| Seat Comfort                       |
| In-Flight Entertainment            |
| Onboard Service                    |
| Leg Room Service                   |
| Baggage Handling                   |
| Check-In Service                   |
| In-Flight Service                  |
| Cleanliness                        |
| ------                             |


<br>

```{r ratings_combine, include = FALSE}

# Select columns for Group1
ratings_group1 <- select(data, Inflight.wifi.service, Departure.Arrival.time.convenient, Ease.of.Online.booking, Gate.location, Online.boarding)  

# Calculate the average for Group1
data$Pre_Flight_and_WiFi_Ratings <- rowMeans(ratings_group1, na.rm = TRUE)

# Select columns for Group2
ratings_group2 <- select(data, Food.and.drink, Seat.comfort, Inflight.entertainment, On.board.service, Leg.room.service, Baggage.handling, Checkin.service, Inflight.service, Cleanliness)

# Calculate the average for Group2
data$In_Flight_and_Baggage_Ratings <- rowMeans(ratings_group2, na.rm = TRUE)
summary(data)

data_ratings_combined <- data[c("Pre_Flight_and_WiFi_Ratings","In_Flight_and_Baggage_Ratings")]


```

```{r ratings_matrix_final, include = FALSE}

cor_matrix_2 <- cor(data_ratings_combined, method = 'spearman')

ratings_correlation_coeff_final <- cor_matrix_2['Pre_Flight_and_WiFi_Ratings','In_Flight_and_Baggage_Ratings']

ratings_correlation_coeff_final

```

<br>

The two consolidated ratings variables share a weak positive correlation (correlation coefficient = **`r ratings_correlation_coeff_final`**), indicating that they can be jointly included in our model without violating the collinearity assumption.

<br>


```{r ratings_final_visual, fig.width = 15, fig.height = 15, fig.align = 'center', echo = FALSE}

# Define ratings subset of main data frame
data_ratings_final <- data[c('Pre_Flight_and_WiFi_Ratings','In_Flight_and_Baggage_Ratings')]

# Transform data
data_long_final <- tidyr::gather(data_ratings_final, variable, value)

# Define custom color palette (necessary for large number of ratings)
colors <- c(
  "red", "blue", "green", "purple", "orange",
  "pink", "gray", "yellow", "cyan", "magenta",
  "brown", "violet", "turquoise", "pink"
)

commas <- function(x) {
  format(x, big.mark = ",")
}

# Create a clustered histogram with color-separated columns
ratingsgraph_final <- ggplot(data_long_final, aes(x = value, fill = variable)) +
  geom_histogram(binwidth = 0.5, position = position_dodge(1), alpha = 0.7) +  # Create the histogram
  labs(x = "\nValue", y = "Frequency\n") +  
  scale_fill_manual(values = colors) +  
  theme(
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold") 
  ) +
  scale_y_continuous(labels = commas) + 
  labs(fill = "Ratings Variables")

ratingsgraph_final + ggtitle("Histogram of aggregated ratings variables\n")
```
<br>


<br>

### EDA: Linearity with log odds

<br>
```{r encode satisfaction}
# Encode the satisfaction variable as 1/0 to use temporarily for JB analysis section, should be removed since this will be closer to the start in the full script

data$satisfaction <- ifelse(data$satisfaction == "satisfied", 1, 0)

# REMOVE IN FINAL
```
<br>
```{r discrete, fig.width = 15, fig.height = 15, fig.align = 'center', echo = FALSE}

# Calculate the discrete buckets for each variable
age_breaks <- seq(0, 90, by = 2)
dist_breaks <- seq(0, 5000, by = 20)
delay_breaks <- seq(0, 1750, by = 10)
rating_breaks <- seq(0,5,by=0.1)

# Use cut() to create breaks
data$AgeCategory <- cut(data$Age, breaks = age_breaks)
data$DistCategory <- cut(data$Flight.Distance, breaks = dist_breaks)
data$DepDelayCategory<- cut(data$Departure.Delay.in.Minutes, breaks = delay_breaks)
data$ArrDelayCategory<- cut(data$Arrival.Delay.in.Minutes, breaks = delay_breaks)
data$PreFlightCategory<-cut(data$Pre_Flight_and_WiFi_Ratings, breaks = rating_breaks)
data$InFlightCategory<-cut(data$In_Flight_and_Baggage_Ratings, breaks = rating_breaks)




```

<br>

Unlike a standard linear regression, which assumes that independent parameters have a linear relationship with the dependent variable, **logistic regression assumes that parameters have a linear relationship with the log odds**.\

Odds represent the number of favorable outcomes divided by the number of unfavorable outcomes. Put differently, if "p" represents the probability of favorable outcomes, Odds = **p/(1-p)**. Log odds take the natural log of the odds, which can be expressed as **ln(p/1-p))**.

We can use a visual test to examine whether or not this assumption holds true for continuous variables. While it is not sensible to compute log odds for individual data points, we can group continuous variables into discrete buckets to examine whether or not they would satisfy this assumption.

<br>

```{r log_odds, include = FALSE}

# Define a function to calculate the log odds for a given x value, to use with aggregate()

log_odds_calc <- function(x) {
  avg <- mean(x)
  log_odds <- log(avg / (1 - avg))
  return(log_odds)
}

# Use aggregate with the log odds function for each continuous variable of interest (with their discrete grouping)
log_odds_Age <- aggregate(satisfaction ~ AgeCategory, data, log_odds_calc)
log_odds_Dist <- aggregate(satisfaction ~ DistCategory, data, log_odds_calc)
log_odds_DepDelay <- aggregate(satisfaction ~ DepDelayCategory, data, log_odds_calc)
log_odds_ArrDelay <- aggregate(satisfaction ~ ArrDelayCategory, data, log_odds_calc)

log_odds_PreFlight <- aggregate(satisfaction ~ PreFlightCategory, data, log_odds_calc)
log_odds_InFlight <- aggregate(satisfaction ~ InFlightCategory, data, log_odds_calc)

```
<br>

```{r log_odds_cont_visual, fig.width = 15, fig.height = 15, fig.align = 'center', echo = FALSE}

# Create scatter plots for "original" continuous variables

ggplot(log_odds_Age, aes(x = AgeCategory, y = satisfaction)) +
  geom_point(size=6, color="turquoise") +
  labs(
    x = "\nAge",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Age\n"
  ) + 
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )  


ggplot(log_odds_Dist, aes(x = DistCategory, y = satisfaction)) +
  geom_point(size=6, color="blue") +
  labs(
    x = "\nDistance",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Distance\n"
  ) +
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )


ggplot(log_odds_DepDelay, aes(x = DepDelayCategory, y = satisfaction)) +
  geom_point(size=6, color="purple") +
  labs(
    x = "\nDeparture Delay",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Departure Delay\n"
  ) + 
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )


ggplot(log_odds_ArrDelay, aes(x = ArrDelayCategory, y = satisfaction)) +
  geom_point(size=6, color="pink") +
  labs(
    x = "\nArrival Delay",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Arrival Delay\n"
  ) +
  
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  ) 

``` 

<br>
<br>
<br>
Based on the graphs, it appears that only flight distance has a roughly linear relationship with log odds of satisfaction. Age appears to have a parabolic relationship, peaking in the middle; some sort of aggressive transformation method may be required to reach a linear relationship. Meanwhile, log odds for both delay statistics quickly disperse in both directions as they increase (likely in part due to the limited frequency of higher durations), making it difficult to conclude with certainty that a linear relationship exists.


```{r log_odds_ratings_visual, fig.width = 15, fig.height = 15, fig.align = 'center', echo = FALSE}

# Create scatter plots for aggregate ratings variables

ggplot(log_odds_PreFlight, aes(x = PreFlightCategory, y = satisfaction)) +
  geom_point(size=6, color="red") +
  labs(
    x = "\nPre-Flight & Wifi Satisfaction Rating",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by Pre-Flight & Wifi Ratings\n"
  ) + 
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )  

ggplot(log_odds_InFlight, aes(x = InFlightCategory, y = satisfaction)) +
  geom_point(size=6, color="orange") +
  labs(
    x = "\nIn-Flight & Baggage Satisfaction Rating",
    y = "Log Odds\n",
    title = "\n\n\nLog Odds by In-Flight and Baggage Ratings\n"
  ) + 
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text.x = element_blank(),
    axis.text = element_text(size = 20),     # Adjust the size of text elements
    axis.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 20, face = "bold"), 
    legend.text = element_text(size = 20),    
    legend.key.size = unit(2, "lines"),
    plot.title = element_text(size = 32, face = "bold")
  )  


```
<br>

In-flight and baggage ratings have a strikingly linear relationship with log odds; meanwhile, pre-flight and wi-fi ratings appear to have a significantly looser connection with a potential dip in log odds for average ratings. We can conclude with confidence that the in-flight aggregate fulfills the linearity assumption, while the pre-flight ratings are far more obscure in that regard. 
